### 2024-08-08 paper.tex
mostly finish, and greatly simplify, 'normalization' argument for jordan basis

\subsection{In the case of general \(X\)}
In the case where \(X\) was a single Jordan block, it was helpful to have the condition that all the nonzero \(a\)'s were to the right of all the nonzero \(b\)'s.
In the general case, it is not clear that such a thing should be true.
And indeed it is not.
However, we will see that we can change basis of \(\mathbb{C}^m\) so that a similar thing is true.
\par Let \(c_{i',j'}^i \in \mathbb{C}\) be some arbitrary coefficients.
Given our Jordan basis \(e_{ij}\) of \(\mathbb{C}^m\), we can define a new Jordan basis \(e_{ij}'\) of \(\mathbb{C}^m\) by 
\[e_{i, \lambda_i}' = e_{i, \lambda_i} + \sum_{\{(i',j') : i' \geq i\} \setminus \{(i, j)\}} c^i_{i',j'} e_{i',j'}.\]
Then we define
\[e_{i,j}' = X^{\lambda_i - j} e_{i, \lambda_i}' = e_{ij} + \sum_{\{(i', j') : i' \geq i\} \setminus \{(i, j)\}} c^i_{i',j'} e_{i', j' - (\lambda_i - j)}.\]
Let us calculate the new value \(a'\) of \(a\) in this new basis.
That is, we want the values \(a_{ij}'\) that satisfy \(Ae_{ij}' = e_{i, j - 1}' + a_{ij}' f_1\).
Looking at the definition of \(e_{ij}'\) above, we see that
\[a_{ij}' = a_{ij} + \sum_{\{(i',j') : i' \geq i\} \setminus \{(i, j)\}} c_{i',j'}^i a_{i', j' - (\lambda_i - j)}.\]
\par Now, for each \(i\), choose the coefficients \(c_{i',j'}^i\) to maximize the value \(\max\{j : (a_{i,1}', ..., a_{i, j}') = 0\}\).
In other words, we choose \(c_{i'j'}^i\) to zero out the largest possible prefix of \((a_{i, 1}, ..., a_{i, \lambda_i})\).
\begin{lemma}
    If the condition of \cref{ab_nilp} on \(a'\) and \(b'\) holds, then there exist nonnegative \(p_i, q_i, r_i\), \(\alpha_{i, j}\), \(\beta_{i, j}\) such that for each \(i\), the following conditions are satisfied.
    \begin{itemize}
        \item \(\lambda_i = p_i + q_i + r_i\)
        \item \((a_{i,1}', ..., a_{i,\lambda_i}') = (0, 0, ..., 0, \alpha_{i,1}, ..., \alpha_{i, r_i})\)
        \item \((b_{i, 1}', ..., b_{i, \lambda_i}') = (\beta_{i, 1}, ..., \beta_{i, p_i}, 0, 0, ..., 0)\)
        \item If \(r_i \neq 0\), then \(\alpha_{i,1} \neq 0\)
        \item If \(p_i \neq 0\), then \(\beta_{i, p_i} \neq 0\)
    \end{itemize}
\end{lemma}
\begin{proof}
    We begin with the first row of \(\lambda\), namely \(i = 1\).
    Let \(k = \min\{j : a'_{1,j} \neq 0\}\) and \(l = \max\{j : b'_{1, j} \neq 0\}\).
    I just need to show that \(l < k\).
    Suppose not.
    Then \((l - k)\) is nonnegative, and by \cref{ab_nilp} we have 
    \[0 = \sum_{ij} b_{ij}' a_{j - (l - k)}' = b_{1,l}' a_{1,k}' + \sum_{i' \geq 1, (i', j') \neq (1, \lambda_1)} b'_{ij} a'_{j - (l - k)}.\]
    This suggests defining 
    \[e_{1,\lambda_1}'' = b_{1, l}'e_{1, \lambda_1}' + \sum_{i' \geq 1, (i', j') \neq (1, \lambda_1)} b'_{i',j' - (\lambda_1 - l)} e'_{i'j'}.\]
    Then for \(j \leq k\),
    \[a_{1,j}'' = b'_{1, l} a'_{1, j} + \sum_{i' \geq 1, (i', j') \neq (1, \lambda_1)} b'_{i', j' - (\lambda_1 - l)} a'_{i', j' - (\lambda_1 - j)} = \]
    \[\sum_{i', j'} b'_{i', j' - (\lambda_1 - l)} a'_{i', j' - (\lambda_1 - j)} = \sum_{i', j'} b'_{i', j'} a'_{i', (j' + (\lambda_1 - l)) - (\lambda_1 - j)} = \sum_{i',j'} b'_{i',j'} a'_{i', j' - (l - j)}.\]
    This last thing is zero by \cref{ab_nilp}, using that \(l - j \geq 0\).
    But this contradicts the definition of the \(c_{i'j'}^1\), because we made it so that \((a''_{1, 1}, ..., a''_{1, k}) = 0\).

>>>>

\subsection{A `normalization' fact about Jordan bases}
Let \(V\) be a finite-dimensional vector space, \(A : V \to V\) a nilpotent operator, and \(f : V \to \mathbb{C}\) a linear map.
Let \((e_{ij} : i \leq m, j \leq \lambda_i)\) be a Jordan basis for \(A\).

\begin{lemma}
    There is a Jordan basis \(e_{ij}\) for \(A\) such that there is at most one pair \((i, j)\) with \(f(e_{ij}) \neq 0\).
\end{lemma}
\begin{proof}
    For any change of basis \(P : V \to V\) commuting with \(A\), we obtain a new Jordan basis \((P(e_{ij}) : i \leq m, j \leq \lambda_i)\).
    For any such \(P\), define 
    \[S_P = \sum_i \begin{cases}-1, & \forall j. \; f \circ P(e_{ij}) = 0 \\ \lambda_i - \min\{j : f \circ P(e_{ij}) \neq 0\}, & \textrm{otherwise} \end{cases}.\]
    Let \(P\) be any operator, among all invertible operators commuting with \(A\), which minimizes \(S_P\).
    Write \(e_{i,j}' := P(e_{i,j})\).
    \par Suppose for contradiction that there are two distinct \(i\)'s (and some \(j\)'s) with \(f(e_{ij}') \neq 0\).
    Then we can take \(e_{i_1, j_1}'\) and \(e_{i_2, j_2}'\), where for \(k \in \{1,2\}\) we have \(f(e_{i_k, j_k}') \neq 0\), and \(\forall j < j_k. \; f(e_{i_k, j}') = 0\).
    Wlog, assume \(\lambda_{i_1} - j_1 \leq \lambda_{i_2} - j_2\).
    Then, we can define \(Q : V \to V\) by
    \begin{itemize}
        \item \(Q(e_{i_1, \lambda_1}') := e_{i_1, \lambda_1}' - \frac{f(e'_{i_1, j_1})}{f(e'_{i_2, j_2})}e_{i_2, j_2 + (\lambda_{i_1} - j_1)}'\)
        \item For \(j < \lambda_1\), \(Q(e_{i_1, j}') := A^{\lambda_{i_1} - j} Q(e_{i_1, \lambda_{i_1}}')\)
        \item For \(i \neq i_1\), \(Q(e_{ij}') = e_{ij}'\).
    \end{itemize}
    Clearly \(Q\) is invertible, and it commutes with \(A\).
    Further, I claim that \(S_{QP}  < S_P\).
    It suffices to show that \(\forall j \leq j_1. \; f \circ Q(e_{i_1,j}') = 0\).
    We have 
    \[f \circ Q(e_{i_1,j}') = f\left(A^{\lambda_{i_1} - j}\left(e_{i_1, \lambda_1}' - \frac{f(e'_{i_1, j_1})}{f(e'_{i_2, j_2})}e_{i_2, j_2 + (\lambda_{i_1} - j_1)}'\right)\right) =\]
    \[f\left(e_{i_1, j}' - \frac{f(e'_{i_1, j_1})}{f(e'_{i_2, j_2})} e_{i_2, j_2 + (j - j_1)}'\right) = f(e_{i_1, j}') - \frac{f(e'_{i_1, j_1})}{f(e'_{i_2, j_2})} e'_{i_2, j_2 + (j - j_1)}.\]
    Clearly (by design), this expression is zero when \(j = j_1\).
    And for \(j < j_1\), we have \(f(e'_{i_1, j}) = f(e'_{i_2, j_2 + (j_2 - j_1)}) = 0\), so it is zero then as well.
    Thus we see that \(S_{QP} < S_P\), contradicting that \(S_P\) is minimal.
    So there must be at most one \(i\) such that there exists \(j\) such that \(f(e_{ij}') \neq 0\).
    \par If there is no such \(i\), we have found the desired basis.
    So, suppose there is such an \(i_0\).
    Let \(U = \langle e_{i_0, j} \rangle\).
    Simply write \(e_j := e_{i_0, j}\).
    For any \(P : U \to U\), let \(S_P := \#\{j : f(P(e_j)) \neq 0\}\).
    Choose \(P\) to minimize \(S_P\), and then let \(e_j' := P(e_j)\).
    Suppose, for contradiction, that \(S_P > 1\).
    Then let \(j_0 = \min\{j : f(e_j') \neq 0\}\), and let \(k\) be distinct from \(j\) with \(f(e_k') \neq 0\).
    Then make the change of basis \(e_{\lambda_{i_0}}'' := e_{\lambda_{i_0}}' + ...\).
\end{proof}

\subsection{In the case of general \(X\)}
In the case where \(X\) was a single Jordan block, it was helpful to have the condition that all the nonzero \(a\)'s were to the right of all the nonzero \(b\)'s.
This allowed some tricks with right-shifting.
In the general case, the parameters \(a\) and \(b\) are not necessarily in such a helpful form.
We will see, however, that we can change basis of \(\mathbb{C}^m\) to put them in a nice form.
